Concurrent Programming with Harmony

### 2020-07-23 book/paper.tex
Updated syntax
Added bitwise operators

\begin{code}
\begin{verbatim}
$ harmony triangle.hny
#states = 13
no issues found
$
\end{verbatim}
\end{code}

@@

In Harmony, constants have a default value,
but those can be overridden on the command
line using the \texttt{-c} option.
\index{constant}%
For example, if you want to test the code for \texttt{N = 100}, run:
\begin{code}
\begin{verbatim}
$ harmony -c N=100 triangle.hny
#states = 103
no issues found
$
\end{verbatim}
\end{code}

@@

\begin{figure}
\begin{code}
\harmonysource{cs}
\end{code}
\caption{\harmonylink{code/csbarebones.hny} Harmony model of a critical section.}
\label{fig:cs}
\end{figure}

@@

Harmony is not an object-oriented language like Python is.  In Python
you can pass a reference to an object to a method, and that method
can then update the object.  In Harmony, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this (as well as some other uses), Harmony supports \emph{pointers}
\index{pointer}%
to shared variables.
If $x$ is a shared variable, then the expression \&$x$
 is a pointer to $x$ (also known as the \emph{address}
\index{address}%
of $x$).
Conversely, if $p$ is a pointer to a shared variable, then the
expression \texttt{\^{}}$p$ is the value of the shared variable.
In particular, \texttt{\^{}\&}$x$ == $x$.
This is similar to how C pointers work.

\autoref{fig:petersonmethods} again shows Peterson's algorithm,
but this time with methods defined to enter and exit the critical
section.
The name \textit{mutex} is often used to denote a variable or value
that is used for mutual exclusion.
\texttt{P\_mutex} is a method that returns a ``mutex,'' which, in this
case, is a dictionary that contains Peterson's Algorithm's state:
a turn variable and two flags.
Both methods \texttt{P\_enter} and \texttt{P\_exit} take two arguments:
a pointer to a mutex and the process identifier (0 or 1).
\texttt{(\^{}}\textit{pm}\texttt{.turn} is the value of the \texttt{.turn} key
in the dictionary that \textit{pm} points to. 
Note that the parentheses in this expression
are needed, as \texttt{\^{}}\textit{pm}\texttt{.turn} would wrongly evaluate to
\texttt{\^{}}(\textit{pm}\texttt{.turn}).

@@

$\mathcal{I}_1$ is an inductive invariant.
To see that invariant $\mathcal{I}_1$ is maintained, note that
\texttt{\^{}}$p$ == \texttt{True} upon entry of \texttt{tas}
(because of the condition on the \textbf{while} loop that the
\texttt{tas} method is invoked in).
So there are two cases:
\begin{enumerate}
\item \texttt{\^{}}$s$ is \texttt{False} upon entry to \texttt{tas}.
Then upon exit \texttt{\^{}}$p$ == \texttt{False} and \texttt{\^{}}$s$ == \texttt{True},
maintaining the invariant.
\item \texttt{\^{}}$s$ is \texttt{True} upon entry to \texttt{tas}.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $\mathcal{I}_1$ is also easy to verify for exiting the critical section.
Invariant $\mathcal{I}_2(i)$ is obvious as (i) process $i$ only proceeds to the critical
section if \textit{private}[$i$] == \texttt{False}, and (ii) no other process modifies
\textit{private}[$i$].

@@

An important invariant of the SBS technique is that the sum of the $N+1$ semaphores
must always be 0 or 1, and it should be 0 when a process is accessing the shared
variables and 1 when not.
As it were, these semaphores taken together are emulating a single binary semaphore.
Initially, \textit{mutex} is~1 and the other $N$ semaphores
are all~0.  To maintain the invariant, each process alternates \texttt{P} and
\texttt{V} operations, starting with \texttt{P}(\&\textit{mutex}) when it tries to
enter the CCS, and ending with a \texttt{V} operation on one of the semaphores
(yet to be determined).

@@

\begin{figure}
\harmonysource{mesa}
\caption{Implementation of condition variables in the \texttt{synch} module.}
\label{fig:cv}
\end{figure}

@@

In \texttt{release\_rlock}, notice that \texttt{notify}(\&\textit{wcond}) is invoked when there
are no readers left, \emph{without} checking if there are writers waiting to enter.
With Mesa monitors this is ok, because calling \texttt{notify} is a no-op if no process
is waiting.

\texttt{release\_wlock} executes \texttt{notifyAll}(\&\textit{wcond}) as well as
\texttt{notify}(\&\textit{wcond)}.
Again, because we do not keep track of the number of waiting readers or writers, we
have to conservatively assume that all waiting readers can enter, or, alternatively,
up to one waiting writer can enter.  So \texttt{release\_wlock} wakes up all
potential candidates.
There are two things to note here.  First, unlike split binary semaphores or Hoare
monitors, where multiple waiting readers would have to be signaled one at a time in a
baton-passing fashion (see \autoref{fig:RWsplitsema}), with Mesa monitors
all readers are awaked in one fell swoop using \texttt{notifyAll}.
Second, both readers and writers are awakened---this is ok because both execute
\texttt{wait} within a \textbf{while} loop, re-checking the condition that they
are waiting for.  So if both type of processes are waiting, either all the readers
get to enter next or one of the writers gets to enter next.

@@

\section*{Exercises}
\begin{problems}
\item Implement a ``try lock'' using Mesa condition variables
(see also Exercise~\ref{ex:trylock}).  It should
have the following API:
\begin{enumerate}
\item \textit{tl} = \texttt{TryLock}() \# \emph{allocate a try lock}
\item \texttt{tlLock}(\&\textit{tl}) \# \emph{acquire a try lock}
\item \texttt{tlTrylock}(\&\textit{tl}) \# \emph{attempt to acquire a try lock}
\item \texttt{tlUnlock}(\&\textit{tl}) \# \emph{release a try lock}
\end{enumerate}
\noindent
\texttt{tlTrylock} should not wait.
Instead it should return \texttt{True} if the lock was successfully
acquired and \texttt{False} if the lock was not available.
\item Implement semaphores using Mesa condition variables.
\item Write a new version of the GPU allocator in \autoref{ex:gpu}
using Mesa condition variables.
In this version,
a process is allowed to allocate a set of GPUs and release a set of GPUs that it
has allocated.  Method \texttt{gpuAllocSet(n)} should block until $n$ GPUs are
available, but it should grant them as soon as they are available.
Method \texttt{gpuReleaseSet(s)} takes a set of GPU identifiers as argument.
A process does not have to return all the GPUs it allocated at once.
\item The specification in the previous question makes the solution unfair.
Explain why this is so.  Then change the specification and the solution so that
it is fair.
\end{problems}

@@

Because of the network's properties, we can model each
channel as a variable that either contains a network message or nothing
(we use \texttt{()} in the model).  Let \textit{s\_chan} be the channel
from Alice to Bob and \textit{r\_chan} the channel from Bob to Alice.
\texttt{net\_send}(\textit{pchan}, $m$, \textit{reliable}) models sending
a message $m$ to
\texttt{\^{}}\textit{pchan}, where \textit{pchan} is either \&\textit{s\_chan}
or \&\textit{r\_chan}.
The method places either $m$ (to model a successful send)
or \texttt{()} (to model loss) in \texttt{\^{}}\textit{pchan}.
The use of the \textit{reliable} flag will be explained later.
\texttt{net\_recv}(\textit{pchan}) models checking \texttt{\^{}pchan} for
the next message.

@@

As a first example, we will revisit the producer/consumer problem.
The code in \autoref{fig:hw} is based on code developed by Herlihy and
Wing~\cite{HW87}.
The code is a ``proof of existence'' for non-blocking synchronization; it
is not necessarily practical.
There are two variables.  \textit{items} is an unbounded array with each
entry initialized to \texttt{()}.  \textit{back} is an index into the
array and points to the next slot where a new value is inserted.
The code uses two interlock instructions:
\begin{itemize}
\item \texttt{inc}($p$): atomically increments \texttt{\^{}}$p$ and returns
the old value;
\item \texttt{swap}($p$): sets \texttt{\^{}}$p$ to \texttt{()} and returns
the old value.
\end{itemize}

Method \texttt{produce}(\textit{item}) uses \texttt{inc}(\&\textit{back})
to allocate
the next available slot in the \textit{items} array.
It stores the item as a singleton tuple.
Method \texttt{consume()} repeatedly scans the array, up to the
\textit{back} index, trying to find an item to return.
To check an entry, it uses \texttt{swap()}
to atomically remove an item from a slot if there is one.
This way, if two or more processes attempt to extract an item from
the same slot, at most one will succeed.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
Boolean & \texttt{False}, \texttt{True} \\
\hline
Integer & \texttt{-inf, ..., -2, -1, 0, 1, 2, ..., inf} \\
\hline
Atom & \texttt{.example}, \texttt{.test1} \\
\hline
Program Counter & (method names are program counter constants) \\
\hline
Dictionary & \texttt{dict\{ .account: 12345, .valid: False \} } \\
\hline
Set & \texttt{\{ 1, 2, 3 \}, \{ False, .id, 3 \} } \\
\hline
Address & \texttt{\&lock, \&flags[2], None} \\
\hline
Context & (generated by \textbf{stop} expression) \\
\hline
\end{tabular}
\vspace{1em}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{$e$ == $e$}, \texttt{$e$ != $e$} & two Harmony values of the same type \\
\hline
\texttt{$e$ < $e$, $e$ <= $e$} & any two Harmony values \\
\hline
\texttt{$e$ > $e$, $e$ >= $e$} & any two Harmony values \\
\hline
\texttt{$e$ and $e$ and ...} & two or more booleans \\
\hline
\texttt{$e$ or $e$ or ...} & two or more booleans \\
\hline
\texttt{not $e$} & a boolean \\
\hline
\texttt{-$e$} & an integer \\
\hline
\texttt{$e$ + $e$ + ...} & two or more integers, sets, or lists \\
\hline
\texttt{$e$ - $e$} & two integers or sets \\
\hline
\texttt{$e$ * $e$ * ...} & two or more integers or sets \\
\hline
\texttt{$e$ / $e$} & two integers (integer division) \\
\hline
\texttt{$e$ \% $e$} & two integers (division remainder) \\
\hline
\texttt{$e$~..~$e$} & two integers \\
\hline
\texttt{$e$ [not] in $e$} & first is any Harmony value, second is set \\
\hline
\texttt{$e$ if $e$ else $e$} & middle $e$ must be a boolean \\
\hline
\texttt{\^{}$e$} & an address \\
\hline
\texttt{choose $e$} & a set \\
\hline
\texttt{min $e$, max $e$} & a set \\
\hline
\texttt{cardinality $e$} & a set \\
\hline
\texttt{keys $e$, len $e$} & a dictionary (includes tuple or list) \\
\hline
\texttt{bagsize $e$} & a bag \\
\hline
\texttt{atLabel $e$} & atom (corresponding to a label) \\
\hline
\texttt{nametag $e$} & $e$ must be \texttt{()} \\
\hline
\texttt{processes $e$} & $e$ must be \texttt{()} \\
\hline
\texttt{setintlevel} \textit{e} & \textit{e} is a boolean \\
\hline
\texttt{stop} \textit{lv} & \textit{lv} is an lvalue \\
\hline
\texttt{\&}\textit{lv} & \textit{lv} is an lvalue \\
\hline
\end{tabular}
\vspace{1em}

\texttt{+} computes the union when applied to sets and
the concatenation when applied to lists or tuples.
\texttt{-} computes set difference when applied to two sets.
\texttt{*} computes the intersection when applied to sets.
\texttt{$x$..$y$} computes the set consisting of the integers $x$
through $y$ inclusive.  It returns the empty set if $x > y$.

An \emph{lvalue}
\index{lvalue}%
(short for left hand value of an assignment statement)
is something that can be assigned.  This can be a
shared variable, a process variable, or a dereferenced pointer
variable.  It can also be indexed.  Examples include
\textit{var}, \textit{var}[$e$], and \texttt{\^{}}\textit{p}.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
Set comprehension & \texttt{\{ f($e$) for $e$ in $s$ \}} \\
\hline
List comprehension & \texttt{[ f($e$) for $e$ in $s$ ]} \\
\hline
Dict comprehension & \texttt{dict\{ f($e$) for $e$ in $s$ \}} \\
\hline
\end{tabular}
\vspace{1em}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{bagEmpty()} & returns an empty bag\\
\hline
\texttt{bagFromSet($s$)} & create a bag from set $s$\\
\hline
\texttt{bagCount($b$, $e$)} & count how many times $e$ occurs in bag $b$\\
\hline
\texttt{bagChoose($b$)} & like \texttt{choose(s)}, but applied to a bag\\
\hline
\texttt{bagAdd}(\textit{pb}, $e$) & add one copy of $e$ to bag \texttt{\^{}}\textit{pb}\\
\hline
\texttt{bagRemove}(\textit{pb}, $e$) & remove one copy of $e$ from bag \texttt{\^{}}\textit{pb}\\
\hline
\end{tabular}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
tas(\textit{lk}) & test-and-set on \texttt{\^{}}\textit{lk} \\
\hline
Lock() & return a lock object \\
\hline
lock(\textit{lk}) & acquire \texttt{\^{}}\textit{lk} \\
\hline
unlock(\textit{lk}) & release \texttt{\^{}}\textit{lk} \\
\hline
Condition(\textit{lk}) & return a condition variable on \texttt{\^{}}\textit{lk} \\
\hline
wait($c$) & wait on condition variable \texttt{\^{}}$c$ \\
\hline
notify($c$) & notify a process waiting on condition variable \texttt{\^{}}$c$ \\
\hline
notifyAll($c$) & notify all processes waiting on condition variable \texttt{\^{}}$c$ \\
\hline
Semaphore(\textit{cnt}) & return a semaphore object initialized to \textit{cnt} \\
\hline
P(\textit{sema}) & procure \texttt{\^{}}\textit{sema}  \\
\hline
V(\textit{sema}) & vacate \texttt{\^{}}\textit{sema}  \\
\hline
Queue() & return a synchronized queue object \\
\hline
dequeue($q$) & return next element of $q$, blocking if empty \\
\hline
enqueue($q$, \textit{item}) & add \textit{item} to $a$ \\
\hline
\end{tabular}

@@

\newpage
Clarifications:
\begin{itemize}
\item The \texttt{Address} instruction expects $n$ values $v_1, ..., v_n$ on the stack.
The top value must be an address value.  The remaining values are keys into dictionaries.
The instruction then computes the address of \texttt{\^}$v_1~v_2~...~v_n$.
\item Even though Harmony code does not allow taking addresses of process variables, both
shared and process variables can have addresses.
\item The \texttt{Load}, \texttt{LoadVar}, \texttt{Del}, \texttt{DelVar},
and \texttt{Stop} instructions have an optional
variable name: if omitted the top of the stack must contain the address of
the variable.
\item \texttt{Store} and \texttt{StoreVar} instructions have an optional
variable name.  In both cases the value to be assigned is on the top
of the stack.  If the name is omitted, the address is underneath that
value on the stack.
\item The effect of the \texttt{Apply} instructions depends much on $m$.
If $m$ is a dictionary, then \texttt{Apply} finds $i$ in the dictionary
and pushes the value.
If $m$ is a program counter, then \texttt{Apply} invokes method $m$ by
pushing the current program counter and setting the program counter to
$m$.  $m$ is supposed to leave the result on the stack.
\item The \texttt{Frame} instruction pushes the value of the process
register (\emph{i.e.}, the values of the process variables) onto the
stack.  It initializes the \texttt{result} variable to the empty dictionary.
The \texttt{Return} instruction restores the process register by popping
its value of the stack.
\item All method calls have exactly one argument, although it sometimes
appears otherwise:
\begin{itemize}
\item \texttt{m()} invokes method \texttt{m} with the empty dictionary \texttt{()} as argument;
\item \texttt{m($a$)} invokes method \texttt{m} with argument $a$;
\item \texttt{m($a$, $b$, $c$)} invokes method \texttt{m} with tuple \texttt{($a$, $b$, $c$)} as argument.
\end{itemize}
The \texttt{Frame} instruction unpacks the argument to the method and places them into process variables by the given names.
\item Every \texttt{Stop} instruction must immediately be followed by a 
\texttt{Continue} instruction.
\end{itemize}

>>>>

\begin{code}
\begin{verbatim}
$ harmony triangle.hny
#states = 13 diameter = 1
no issues found
$
\end{verbatim}
\end{code}

@@

In Harmony, constants have a default value,
but those can be overridden on the command
line using the \texttt{-c} option.
\index{constant}%
For example, if you want to test the code for \texttt{N = 100}, run:
\begin{code}
\begin{verbatim}
$ harmony -c N=100 triangle.hny
#states = 103 diameter = 1
no issues found
$
\end{verbatim}
\end{code}

@@

\begin{figure}
\begin{code}
\harmonysource{cs}
\end{code}
\caption{\harmonylink{code/cs.hny} Harmony model of a critical section.}
\label{fig:cs}
\end{figure}

@@

Harmony is not an object-oriented language like Python is.  In Python
you can pass a reference to an object to a method, and that method
can then update the object.  In Harmony, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this (as well as some other uses), Harmony supports \emph{pointers}
\index{pointer}%
to shared variables.
If $x$ is a shared variable, then the expression ?$x$
 is a pointer to $x$ (also known as the \emph{address}
\index{address}%
of $x$).
Conversely, if $p$ is a pointer to a shared variable, then the
expression \texttt{!{}}$p$ is the value of the shared variable.
In particular, \texttt{!{}?}$x$ == $x$.
This is similar to how C pointers work.

\autoref{fig:petersonmethods} again shows Peterson's algorithm,
but this time with methods defined to enter and exit the critical
section.
The name \textit{mutex} is often used to denote a variable or value
that is used for mutual exclusion.
\texttt{P\_mutex} is a method that returns a ``mutex,'' which, in this
case, is a dictionary that contains Peterson's Algorithm's state:
a turn variable and two flags.
Both methods \texttt{P\_enter} and \texttt{P\_exit} take two arguments:
a pointer to a mutex and the process identifier (0 or 1).
\texttt{(!{}}\textit{pm}\texttt{.turn} is the value of the \texttt{.turn} key
in the dictionary that \textit{pm} points to. 
Note that the parentheses in this expression
are needed, as \texttt{!{}}\textit{pm}\texttt{.turn} would wrongly evaluate to
\texttt{!{}}(\textit{pm}\texttt{.turn}).

@@

$\mathcal{I}_1$ is an inductive invariant.
To see that invariant $\mathcal{I}_1$ is maintained, note that
\texttt{!{}}$p$ == \texttt{True} upon entry of \texttt{tas}
(because of the condition on the \textbf{while} loop that the
\texttt{tas} method is invoked in).
So there are two cases:
\begin{enumerate}
\item \texttt{!{}}$s$ is \texttt{False} upon entry to \texttt{tas}.
Then upon exit \texttt{!{}}$p$ == \texttt{False} and \texttt{!{}}$s$ == \texttt{True},
maintaining the invariant.
\item \texttt{!{}}$s$ is \texttt{True} upon entry to \texttt{tas}.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $\mathcal{I}_1$ is also easy to verify for exiting the critical section.
Invariant $\mathcal{I}_2(i)$ is obvious as (i) process $i$ only proceeds to the critical
section if \textit{private}[$i$] == \texttt{False}, and (ii) no other process modifies
\textit{private}[$i$].

@@

An important invariant of the SBS technique is that the sum of the $N+1$ semaphores
must always be 0 or 1, and it should be 0 when a process is accessing the shared
variables and 1 when not.
As it were, these semaphores taken together are emulating a single binary semaphore.
Initially, \textit{mutex} is~1 and the other $N$ semaphores
are all~0.  To maintain the invariant, each process alternates \texttt{P} and
\texttt{V} operations, starting with \texttt{P}(?\textit{mutex}) when it tries to
enter the CCS, and ending with a \texttt{V} operation on one of the semaphores
(yet to be determined).

@@

\begin{figure}
\begin{code}
\harmonysource{mesa}
\end{code}
\caption{Implementation of condition variables in the \texttt{synch} module.}
\label{fig:cv}
\end{figure}

@@

In \texttt{release\_rlock}, notice that \texttt{notify}(?\textit{wcond}) is invoked when there
are no readers left, \emph{without} checking if there are writers waiting to enter.
With Mesa monitors this is ok, because calling \texttt{notify} is a no-op if no process
is waiting.

\texttt{release\_wlock} executes \texttt{notifyAll}(?\textit{wcond}) as well as
\texttt{notify}(?\textit{wcond)}.
Again, because we do not keep track of the number of waiting readers or writers, we
have to conservatively assume that all waiting readers can enter, or, alternatively,
up to one waiting writer can enter.  So \texttt{release\_wlock} wakes up all
potential candidates.
There are two things to note here.  First, unlike split binary semaphores or Hoare
monitors, where multiple waiting readers would have to be signaled one at a time in a
baton-passing fashion (see \autoref{fig:RWsplitsema}), with Mesa monitors
all readers are awaked in one fell swoop using \texttt{notifyAll}.
Second, both readers and writers are awakened---this is ok because both execute
\texttt{wait} within a \textbf{while} loop, re-checking the condition that they
are waiting for.  So if both type of processes are waiting, either all the readers
get to enter next or one of the writers gets to enter next.

@@

\section*{Exercises}
\begin{problems}
\item Implement a ``try lock'' using Mesa condition variables
(see also Exercise~\ref{ex:trylock}).  It should
have the following API:
\begin{enumerate}
\item \textit{tl} = \texttt{TryLock}() \# \emph{allocate a try lock}
\item \texttt{tlLock}(?\textit{tl}) \# \emph{acquire a try lock}
\item \texttt{tlTrylock}(?\textit{tl}) \# \emph{attempt to acquire a try lock}
\item \texttt{tlUnlock}(?\textit{tl}) \# \emph{release a try lock}
\end{enumerate}
\noindent
\texttt{tlTrylock} should not wait.
Instead it should return \texttt{True} if the lock was successfully
acquired and \texttt{False} if the lock was not available.
\item Implement semaphores using Mesa condition variables.
\item Write a new version of the GPU allocator in \autoref{ex:gpu}
using Mesa condition variables.
In this version,
a process is allowed to allocate a set of GPUs and release a set of GPUs that it
has allocated.  Method \texttt{gpuAllocSet(n)} should block until $n$ GPUs are
available, but it should grant them as soon as they are available.
Method \texttt{gpuReleaseSet(s)} takes a set of GPU identifiers as argument.
A process does not have to return all the GPUs it allocated at once.
\item The specification in the previous question makes the solution unfair.
Explain why this is so.  Then change the specification and the solution so that
it is fair.
\end{problems}

@@

Because of the network's properties, we can model each
channel as a variable that either contains a network message or nothing
(we use \texttt{()} in the model).  Let \textit{s\_chan} be the channel
from Alice to Bob and \textit{r\_chan} the channel from Bob to Alice.
\texttt{net\_send}(\textit{pchan}, $m$, \textit{reliable}) models sending
a message $m$ to
\texttt{!{}}\textit{pchan}, where \textit{pchan} is either ?\textit{s\_chan}
or ?\textit{r\_chan}.
The method places either $m$ (to model a successful send)
or \texttt{()} (to model loss) in \texttt{!{}}\textit{pchan}.
The use of the \textit{reliable} flag will be explained later.
\texttt{net\_recv}(\textit{pchan}) models checking \texttt{!{}pchan} for
the next message.

@@

As a first example, we will revisit the producer/consumer problem.
The code in \autoref{fig:hw} is based on code developed by Herlihy and
Wing~\cite{HW87}.
The code is a ``proof of existence'' for non-blocking synchronization; it
is not necessarily practical.
There are two variables.  \textit{items} is an unbounded array with each
entry initialized to \texttt{()}.  \textit{back} is an index into the
array and points to the next slot where a new value is inserted.
The code uses two interlock instructions:
\begin{itemize}
\item \texttt{inc}($p$): atomically increments \texttt{!{}}$p$ and returns
the old value;
\item \texttt{swap}($p$): sets \texttt{!{}}$p$ to \texttt{()} and returns
the old value.
\end{itemize}

Method \texttt{produce}(\textit{item}) uses \texttt{inc}(?\textit{back})
to allocate
the next available slot in the \textit{items} array.
It stores the item as a singleton tuple.
Method \texttt{consume()} repeatedly scans the array, up to the
\textit{back} index, trying to find an item to return.
To check an entry, it uses \texttt{swap()}
to atomically remove an item from a slot if there is one.
This way, if two or more processes attempt to extract an item from
the same slot, at most one will succeed.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
Boolean & \texttt{False}, \texttt{True} \\
\hline
Integer & \texttt{-inf, ..., -2, -1, 0, 1, 2, ..., inf} \\
\hline
Atom & \texttt{.example}, \texttt{.test1} \\
\hline
Program Counter & (method names are program counter constants) \\
\hline
Dictionary & \texttt{dict\{ .account: 12345, .valid: False \} } \\
\hline
Set & \texttt{\{ 1, 2, 3 \}, \{ False, .id, 3 \} } \\
\hline
Address & \texttt{?lock, ?flags[2], None} \\
\hline
Context & (generated by \textbf{stop} expression) \\
\hline
\end{tabular}
\vspace{1em}

@@

{\small
\begin{tabular}{|l|l|}
\hline
\texttt{$e$ == $e$}, \texttt{$e$ != $e$} & two Harmony values of the same type \\
\hline
\texttt{$e$ < $e$, $e$ <= $e$} & any two Harmony values \\
\hline
\texttt{$e$ > $e$, $e$ >= $e$} & any two Harmony values \\
\hline
\texttt{$e$ and $e$ and ...} & two or more booleans \\
\hline
\texttt{$e$ or $e$ or ...} & two or more booleans \\
\hline
\texttt{not $e$} & a boolean \\
\hline
\texttt{-$e$, \string~$e$} & an integer \\
\hline
\texttt{$e$ + $e$ + ...} & two or more integers, tuples, or lists \\
\hline
\texttt{$e$ - $e$} & two integers or sets \\
\hline
\texttt{$e$ * $e$ * ...} & two or more integers \\
\hline
\texttt{$e$ \& $e$ \& * ...} & two or more integers or sets \\
\hline
\texttt{$e$ | $e$ | * ...} & two or more integers or sets \\
\hline
\texttt{$e$ \string^ $e$ \string^ * ...} & two or more integers or sets \\
\hline
\texttt{$e$ / $e$} & two integers (integer division) \\
\hline
\texttt{$e$ \% $e$} & two integers (division remainder) \\
\hline
\texttt{$\{ e..e \}$} & two integers \\
\hline
\texttt{$e$ [not] in $e$} & first is any Harmony value, second is set \\
\hline
\texttt{$e$ if $e$ else $e$} & middle $e$ must be a boolean \\
\hline
\texttt{!{}$e$} & an address \\
\hline
\texttt{choose $e$} & a set \\
\hline
\texttt{min $e$, max $e$} & a set \\
\hline
\texttt{cardinality $e$} & a set \\
\hline
\texttt{keys $e$, len $e$} & a dictionary (includes tuple or list) \\
\hline
\texttt{bagsize $e$} & a bag \\
\hline
\texttt{atLabel $e$} & atom (corresponding to a label) \\
\hline
\texttt{nametag $e$} & $e$ must be \texttt{()} \\
\hline
\texttt{processes $e$} & $e$ must be \texttt{()} \\
\hline
\texttt{setintlevel} \textit{e} & \textit{e} is a boolean \\
\hline
\texttt{stop} \textit{lv} & \textit{lv} is an lvalue \\
\hline
\texttt{?}\textit{lv} & \textit{lv} is an lvalue \\
\hline
\end{tabular}
}
\vspace{1em}

\texttt{+} concatenates when applied to lists or tuples.
\texttt{-} computes set difference when applied to two sets.
\texttt{\&} computes the intersection when applied to sets.
\texttt{|} computes the union when applied to sets.
\texttt{\^} computes $(s \cup t) \backslash (s \cap t)$ when applied to sets $s$ and $t$.
\texttt{$\{ x .. y \}$} computes the set consisting of the integers $x$
through $y$ inclusive.  It returns the empty set if $x > y$.

An \emph{lvalue}
\index{lvalue}%
(short for left hand value of an assignment statement)
is something that can be assigned.  This can be a
shared variable, a process variable, or a dereferenced pointer
variable.  It can also be indexed.  Examples include
\textit{var}, \textit{var}[$e$], and \texttt{!{}}\textit{p}.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
Set comprehension & \texttt{\{ f($v$) for $v$ in $s$ \}} \\
\hline
List comprehension & \texttt{[ f($v$) for $v$ in $s$ ]} \\
\hline
Dict comprehension & \texttt{dict\{ f($v$) for $v$ in $s$ \}} \\
\hline
\end{tabular}
\vspace{1em}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{bagEmpty()} & returns an empty bag\\
\hline
\texttt{bagFromSet($s$)} & create a bag from set $s$\\
\hline
\texttt{bagCount($b$, $e$)} & count how many times $e$ occurs in bag $b$\\
\hline
\texttt{bagChoose($b$)} & like \texttt{choose(s)}, but applied to a bag\\
\hline
\texttt{bagAdd}(\textit{pb}, $e$) & add one copy of $e$ to bag \texttt{!{}}\textit{pb}\\
\hline
\texttt{bagRemove}(\textit{pb}, $e$) & remove one copy of $e$ from bag \texttt{!{}}\textit{pb}\\
\hline
\end{tabular}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
tas(\textit{lk}) & test-and-set on \texttt{!{}}\textit{lk} \\
\hline
Lock() & return a lock object \\
\hline
lock(\textit{lk}) & acquire \texttt{!{}}\textit{lk} \\
\hline
unlock(\textit{lk}) & release \texttt{!{}}\textit{lk} \\
\hline
Condition(\textit{lk}) & return a condition variable on \texttt{!{}}\textit{lk} \\
\hline
wait($c$) & wait on condition variable \texttt{!{}}$c$ \\
\hline
notify($c$) & notify a process waiting on condition variable \texttt{!{}}$c$ \\
\hline
notifyAll($c$) & notify all processes waiting on condition variable \texttt{!{}}$c$ \\
\hline
Semaphore(\textit{cnt}) & return a semaphore object initialized to \textit{cnt} \\
\hline
P(\textit{sema}) & procure \texttt{!{}}\textit{sema}  \\
\hline
V(\textit{sema}) & vacate \texttt{!{}}\textit{sema}  \\
\hline
Queue() & return a synchronized queue object \\
\hline
dequeue($q$) & return next element of $q$, blocking if empty \\
\hline
enqueue($q$, \textit{item}) & add \textit{item} to $a$ \\
\hline
\end{tabular}

@@

\newpage
Clarifications:
\begin{itemize}
\item The \texttt{Address} instruction expects $n$ values $v_1, ..., v_n$ on the stack.
The top value must be an address value.  The remaining values are keys into dictionaries.
The instruction then computes the address of \texttt{!}$v_1~v_2~...~v_n$.
\item Even though Harmony code does not allow taking addresses of process variables, both
shared and process variables can have addresses.
\item The \texttt{Load}, \texttt{LoadVar}, \texttt{Del}, \texttt{DelVar},
and \texttt{Stop} instructions have an optional
variable name: if omitted the top of the stack must contain the address of
the variable.
\item \texttt{Store} and \texttt{StoreVar} instructions have an optional
variable name.  In both cases the value to be assigned is on the top
of the stack.  If the name is omitted, the address is underneath that
value on the stack.
\item The effect of the \texttt{Apply} instructions depends much on $m$.
If $m$ is a dictionary, then \texttt{Apply} finds $i$ in the dictionary
and pushes the value.
If $m$ is a program counter, then \texttt{Apply} invokes method $m$ by
pushing the current program counter and setting the program counter to
$m$.  $m$ is supposed to leave the result on the stack.
\item The \texttt{Frame} instruction pushes the value of the process
register (\emph{i.e.}, the values of the process variables) onto the
stack.  It initializes the \texttt{result} variable to the empty dictionary.
The \texttt{Return} instruction restores the process register by popping
its value of the stack.
\item All method calls have exactly one argument, although it sometimes
appears otherwise:
\begin{itemize}
\item \texttt{m()} invokes method \texttt{m} with the empty dictionary \texttt{()} as argument;
\item \texttt{m($a$)} invokes method \texttt{m} with argument $a$;
\item \texttt{m($a$, $b$, $c$)} invokes method \texttt{m} with tuple \texttt{($a$, $b$, $c$)} as argument.
\end{itemize}
The \texttt{Frame} instruction unpacks the argument to the method and places them into process variables by the given names.
\item Every \texttt{Stop} instruction must immediately be followed by a 
\texttt{Continue} instruction.
\end{itemize}
